Standardized crypto-loans on\\ the Cardano blockchain

### 2021-02-03 papers/actus2021/article.tex
Revised sectinos 2-4 and bibliography.

\textbf{DONE:}
\begin{enumerate}
\item
Added overview at the end of the intro.
\item
Tidying up identifiers in math equations using the command \verb+\ident{...}+
\item
Biblio sorted out ordering (added keys). shortened author and editor lists
\end{enumerate}

\textbf{TO DO:}
\begin{enumerate}
\item
Algorithmic Contract Types Unified Standards (ACTUS) -- for first use. Also reference.
\item
Do we include Alex as co-author?
\item
What about R (shiny)?
\end{enumerate}

@@

This section introduces the Marlowe DSL and the ACTUS standard for financial contracts.

@@

Marlowe is a high-level, domain-specific language (DSL) for writing
financial contracts on blockchain\cite{marlowe}. Marlowe is defined
by an executable semantics in Haskell, and has been implemented on
the UTxO-based Cardano blockchain.

The main advantage of using Marlowe to carry ACTUS logic is enhanced
security that static analysis provides. SMT-solvers gain increasing
popularity as an alternative to fuzzing when it comes to testing software
for vulnerabilities\cite{smt}. Smart contracts generated from ACTUS
specification are quite complicated, so manually testing every possible
execution path for every possible combination of contract terms is
not feasible. Static analysis, in turn, can reduce the amount of ``dumb''
mistakes that could cost millions to the users by ensuring obvious
properties of a contract. It is not a panacea, however, unlike with
fuzzing, if SA says that certain property holds - it does it with
100\% assurance.

\subsection{ACTUS}

ACTUS defines the logic embedded in legal agreements that eventually
turn the contract terms into actual cash flows, or more generally
business events \cite{actus}. Most of its basic contract types represent
different variations of lending contracts.

ACTUS relies on a state machine formalism in order to describe the
behaviour of a given contract. Every payoff (transfer of funds in
or out of a contract) can be inferred for any given state. Every state
can be inferred from previous events and observed risk factors:

\noindent 
\begin{equation}
\mbox{\emph{payoff}}_{i}=POF(state_{i})
\end{equation}

\noindent 
\begin{equation}
path=STF_{1}(ct,ev_{1})\circ STF_{2}(ct,ev_{2}).\circ\ldots\circ STF_{i}(ct,ev_{n})
\end{equation}

\noindent 
\begin{equation}
state_{i}=path(INIT(ct)),
\end{equation}

where $ct$ is contract terms, $INIT$ returns initial state, $sched$
returns scheduled events, $STF$ takes ${contractterms,event,state}$
and returns next state, $POF$ returns payoff.

\subsection{Necessity for Oracles}

In order to support variable interest rates and scaling, ACTUS requires
a smart contract to be able to observe the value of a given risk factor
(such as interest rate) at a particular point in time t:

\noindent 
\begin{equation}
\mbox{\emph{riskfactor}}_{i}{}_{t}=O_{rf}(i,t)
\end{equation}

This is due to the risk factor's state not being known at instantiation
time. In the case of Cardano blockchain, these values are usually
provided through an ``Oracle'' mechanism\cite{oracles}. Oracle
could be a trusted party providing necessary data or network of parties
under consensus {[}find a quote{]}. From Marlowe DSL perspective,
the exact mechanism that provides external data is less important,
as Marlowe abstracts over IO by requiring particular type of input
(``Choice'') protected with a cryptographic signature (Choice owner).
As a result, the event of receiving data from Oracle is treated the
same as receiving integer numbers from user input in other languages

@@

\subsection{Mimicking the specification with Haskell code}

The ACTUS standard is specified in terms of scheduling, payoff and
state transition functions that are polymorphic on event and contract
type. The specification also follows quite specific naming conventions
that are incompatible with Haskell's conventions. The executable specification
follows original ACTUS conventions as close as possible in order to
ease code base maintenance when faced with updates of original ACTUS
spec (link to github).

Using Haskell itself as a DSL for explicitly encoding formulas without
using advanced language idioms also simplifies code generation, but
in case of ACTUS comes at a cost reduced type-safety, handling nullable
values explicitly introduces risk of exceptions. However this risk
is addressed using QuickCheck generators.

@@

In order to keep our executable specification independent of the carrier
(whether its smart-contract engine, proof assistant, analytical framework
or even machine learning model) - we abstract over underlying representations
of state variables and arithmetic operations:

\begin{verbatim}
-- Definitions/ContractState.hs 
data ContractStatePoly a b = ContractStatePoly  
{  
tmd     :: b  
, nt    :: a  
, ipnr  :: a  
, ipac  :: a  
, feac  :: a  
, fac   :: a  
, nsc   :: a  
, isc   :: a  
, prf   :: ContractStatus  
, sd    :: b  
, prnxt :: a  
, ipcb  :: a  
} deriving (Show)

-- Ops.hs
class ActusOps a where    
   _min :: a -> a -> a
   _max :: a -> a -> a
   _zero :: a
   _one :: a

@@

class RoleSignOps a where
   _r :: ContractRole -> a
\end{verbatim}

Thus, every formula in the executable spec could either be instantiated
to: 
\begin{itemize}
\item formula on some ``atomic'' type (like Double, Day) - could be used
to directly compute cash-flows for analytical purposes or precompute
payoffs for smart contracts that don't depend on oracles 
\item formula representing piece of AST (Marlowe Value, Observation) - could
be used to generate smart contracts that depend on oracles or to generate
code in another language (like Agda) 
\end{itemize}
This approach of abstracting formulas has a limitation of not allowing
conditionals to be expressed in an abstract way (There is no ActusIf
typeclass) - luckily most of conditional expressions in ACTUS specification
don't depend on variable state of a contract - they depend on ContractTerms
that are known in advance during contract generation. This allows
us to dispatch appropriate formulas during generation rather than
execution.

The only exception to this are rare situations where we need to compare
2 state variables and either choose formula1 or 0 depending on comparison
result:

\begin{equation}
formula(st)=\begin{cases}
formula'(st) & var1(st)<var2(st)\\
0 & otherwise
\end{cases}
\end{equation}

We rely on pseudo-boolean ``less than'' function in order to address
that:

\noindent 
\begin{equation}
formula(st)=pseudoLt(var1(s1),var2(st)))*formula'
\end{equation}

\noindent 
\begin{equation}
pseudoLt(a,b)=Cond(a>b,1,0)
\end{equation}

@@

In order to simplify serialisation/deserialisation of contract terms
across Actus related services maintained by Cardano - we rely on ``superposed''
representation of contract terms - all ACTUS contract types are represented
with the same type. While such representation allows both encoder
and decoder to express any ACTUS contract terms, it also allows for
invalid combinations of terms (for example PRNXT cannot be applied
to PAM-contract), which means contracts require specific validation
that is implemented by means of applicability function:

\noindent 
\begin{equation}
Applicability:ContractTerms\mapsto Bool
\end{equation}

ACTUS defines a family of applicability functions polymorphic on contract
type (TODO link to ACTUS applicability):

\noindent 
\begin{equation}
Applicability:ContractType\times ContractTerm\mapsto ApplicabilityType,
\end{equation}

where applicability could be: none, always, nullable, multiple

In order to build superposed contract terms type for such function,
we have to resolve conflicting applicability for contract terms merged
using the following resolution rules:

\noindent 
\begin{equation}
weaken(a_{1},a_{2})=\begin{cases}
nullable & a_{1}=none\wedge a_{2}=always\\
nullable & a_{1}=always\wedge a_{2}=none\\
a_{1} & priority(a_{1})>priority(a_{2})\\
a_{2} & otherwise
\end{cases}
\end{equation}

\noindent 
\begin{equation}
priority(x)=\begin{cases}
0 & x=always\\
1 & x=none\\
2 & x=nullable\\
3 & x=multiple
\end{cases}
\end{equation}

Where $a_{1}$ - is applicability of a given term of first contract
type to be merged $a_{2}$ - is applicability of a given term of second
contract type to be merged

>>>>

\textbf{DONE:}
\begin{enumerate}
\item
Added overview at the end of the intro.
\item
Tidying up identifiers in math equations using the command \verb+\ident{...}+
\item
Biblio sorted out ordering (added keys). shortened author and editor lists
\item
Revised sections 2--4.
\end{enumerate}

\textbf{TO DO:}
\begin{enumerate}
\item
Deal with inline comments in source code: \verb+% SJT ...+
\item
Write introduction [SJT]
\item
Should we include Alex as co-author?
\item
What about R (shiny)?
\end{enumerate}

@@

This section introduces the Marlowe domain-specific language (DSL) and the ACTUS standard for financial contracts.

@@

Marlowe~\cite{marlowe} is a high-level DSL for writing
financial contracts on blockchain. Marlowe is defined
by an executable semantics in Haskell, and has been implemented on
the UTxO-based Cardano blockchain. 
% SJT added this
Marlowe contracts are finite, and moreover have defined finite lifetime. This means that it is possible to perform \emph{static analysis} on them. For example, it is possible symbolically to check \emph{all} execution paths without running a contract, and in particular checking that every \texttt{Pay} construct can be executed in full.

The main advantage of using Marlowe to carry ACTUS logic is enhanced
security that such a static analysis provides. SMT-solvers~\cite{smt} are gaining
popularity as an alternative to fuzzing when it comes to testing software
for vulnerabilities. Smart contracts generated from ACTUS
specification are quite complicated, so manually testing every possible
execution path for every possible combination of contract terms is
not feasible. Static analysis, in turn, can reduce the amount of ``dumb''
mistakes that could cost millions to the users by ensuring obvious
properties of a contract. It is not a panacea, however, unlike with
fuzzing, if SA says that certain property holds - it does it with
100\% assurance.

\subsection{ACTUS}
\label{ACTUS}

The Algorithmic Contract Types Unified Standards (ACTUS)~\cite{actus} define the logic embedded in legal agreements that eventually
turn the contract terms into actual cash flows, or more generally
business events. Most of its basic contract types represent
different variations of lending contracts.

ACTUS relies on a state machine formalism in order to describe the
behaviour of a given contract. Every \emph{payoff} -- i.e. transfer of funds in
or out of a contract -- can be inferred for any given state. Every state
can be derived from previous events and observed risk factors:

\noindent 
\begin{equation}
\ident{payoff}_{i}=\ident{POF}(\ident{state}_{i})
\end{equation}

\newcommand{\STF}{\ident{STF}}
\noindent 
\begin{equation}
\ident{path}=\STF_{1}(ct,ev_{1})\circ \STF_{2}(ct,ev_{2}).\circ\ldots\circ \STF_{i}(ct,ev_{n})
\end{equation}

\noindent 
\begin{equation}
\ident{state}_{i}=\ident{path}(\ident{INIT}(ct)),
\end{equation}
\noindent
where \emph{ct} stands for contract terms, \emph{INIT} returns initial state, \emph{sched}
returns scheduled events, \emph{STF} takes \emph{contractterms}, \emph{event}, and \emph{state}
and returns the next state, and \emph{POF} returns the payoff in a state.

\subsection{Oracles}

In order to support variable interest rates and scaling, ACTUS requires
a smart contract to be able to observe the value of a given risk factor,
such as an interest rate, at a particular point in time $t$. This is due to the state of the risk factor not being known at instantiation
time.

% SJT - the "i" and "t" subscripts here look a bit odd: is that what's intended?
\noindent 
\begin{equation}
\ident{riskfactor}_{i}{}_{t}=O_{\footnotesize\ident{rf}}(i,t)
\end{equation}
\noindent
In the case of the Cardano blockchain, these values are usually
provided through an \emph{oracle} mechanism\cite{oracles}. An oracle
could be a trusted party providing necessary data or network of parties
under consensus {[}find a reference{]}. 
% SJT REF needed
From a Marlowe DSL perspective,
the exact mechanism that provides external data is less important,
as Marlowe abstracts over IO by requiring a particular type of input -- a 
\texttt{Choice} -- that is protected with a cryptographic signature by the source of the choice.
% SJT changed from "owner" to "source".
As a result, the event of receiving data from an oracle is treated the
same as receiving numeric input in other languages

@@

ACTUS is defined in a textual specification\footnote{Available from \url{https://www.actusfrf.org/techspecs}.} which, while expressed in mathematical notation, is essentially informal. In this section we describe how this specification is turned into an \emph{executable} version by rendering it in Haskell. This translation is in fact a transliteration, since notation, variable names and so forth are respected.
% SJT: are you happy with this? If not, let's rephrase,
% but it would be good include something to this effect.

\subsection{Rendering the specification in Haskell}

The ACTUS standard is specified in terms of scheduling, payoff and
state transition functions that are polymorphic on event and contract
type, as noted in Section \ref{ACTUS} above. The specification also follows quite specific naming conventions
that are incompatible with Haskell's conventions. The executable specification
follows original ACTUS conventions as closely as possible in order to
ease code base maintenance when faced with updates of the ACTUS
spec repository\footnote{\url{https://github.com/actusfrf/actus-techspecs}}.

Using Haskell itself as a DSL for explicitly encoding formulas without
using advanced language idioms also simplifies code generation. 
In case of ACTUS this comes at a cost reduced type-safety, handling nullable
values explicitly introduces risk of exceptions. However this risk
is addressed using property-based testing, and in particular QuickCheck generators. This is discussed in more detail in Section \ref{verification} below.
% SJT Added x-ref to section 7.

@@

In order to keep our executable specification independent of the carrier -- whether it is a smart-contract engine, proof assistant, analytical framework
or even machine learning model -- we abstract over the underlying representation
of state variables,

\begin{verbatim}
-- Definitions/ContractState.hs 
data ContractStatePoly a b = ContractStatePoly  
{  
  tmd     :: b  
  , nt    :: a  
  , ipnr  :: a  
  , ipac  :: a  
  , feac  :: a  
  , fac   :: a  
  , nsc   :: a  
  , isc   :: a  
  , prf   :: ContractStatus  
  , sd    :: b  
  , prnxt :: a  
  , ipcb  :: a  
} deriving (Show) 
\end{verbatim}
\noindent
and arithmetic operations,
\begin{verbatim}
-- Ops.hs
class ActusOps a where    
   _min :: a -> a -> a
   _max :: a -> a -> a
   _zero :: a
   _one :: a

@@

class RoleSignOps a where
   _r :: ContractRole -> a
\end{verbatim}
\noindent
Thus, every formula in the executable spec could be instantiated
to: 
\begin{itemize}
\item a formula on some \emph{atomic} type, like \texttt{Double} or \texttt{Day}, which could be used
to directly compute cash-flows for analytical purposes or precompute
payoffs for smart contracts that do not depend on oracles; or 
\item a formula representing a piece of \emph{abstract syntax}, e.g. a Marlowe \texttt{Value} or \texttt{Observation}, that could
be used to generate smart contracts that depend on oracles or to generate
code in another language, such as Agda.
\end{itemize}
This approach of abstracting formulas has a limitation of not allowing
conditionals to be expressed in an abstract way: in other words, there is no \texttt{ActusIf}
typeclass. Luckily most of conditional expressions in ACTUS specification
don't depend on variable state of a contract, they depend on \texttt{ContractTerms}
that are known in advance during contract generation. This allows
us to dispatch appropriate formulas during generation rather than
execution.

The only exception to this are the rare situations where we need to compare
2 state variables and choose either $\ident{formula}^\prime$ or 0 depending on the result of the comparison
result:

\begin{equation}
\ident{formula}(st)=\begin{cases}
\ident{formula}^\prime(st) & \ident{var}_1(st)<\ident{var}_2(st)\\
0 & \ident{otherwise}
\end{cases}
\end{equation}
We rely on a pseudo-Boolean \emph{less than} function in order to address
that:

\noindent 
\begin{equation}
\ident{formula}(st)=\ident{pseudoLt}(\ident{var}_1(st),\ident{var}_2(st)))*\ident{formula}^\prime
\end{equation}

\noindent 
\begin{equation}
\ident{pseudoLt}(a,b)=\ident{Cond}(a>b,1,0)
\end{equation}

@@

In order to simplify serialisation and deserialisation of contract terms
across ACTUS related services maintained by Cardano we rely on ``superposed''
representation of contract terms: all ACTUS contract types are represented
with the same type.

While such a representation allows both encoder
and decoder to express any ACTUS contract terms, it also allows for
invalid combinations of terms (for example \emph{PRNXT} cannot be applied
to a PAM contract), which means contracts require specific validation
that is implemented by means of the applicability function:

% SJT - do you mean \mapsto or \rightarrow here?
\noindent 
\begin{equation}
\ident{Applicability}:\ident{ContractTerms}\mapsto \ident{Bool}
\end{equation}

ACTUS defines a family of applicability functions polymorphic on contract
type (TODO link to ACTUS applicability):
% SJT something needed here.

\smallskip
% SJT changed ContractTerm to ContractTerms
\noindent 
\begin{equation}
\ident{Applicability} : \ident{ContractType} \times \ident{ContractTerms} \mapsto \ident{ApplicabilityType}
\end{equation}
\noindent
where applicability could be: \emph{none}, \emph{always}, \emph{nullable}, or \emph{multiple}.

In order to build superposed contract terms type for such functions,
we have to resolve conflicting applicability types for merged contract terms
using the following resolution rules:

\noindent 
\begin{equation}
\ident{weaken}(a_{1},a_{2})=\begin{cases}
\ident{nullable} & a_{1}=\ident{none}\wedge a_{2}=\ident{always}\\
\ident{nullable} & a_{1}=\ident{always}\wedge a_{2}=\ident{none}\\
a_{1} & \ident{priority}(a_{1})>\ident{priority}(a_{2})\\
a_{2} & \ident{otherwise}
\end{cases}
\end{equation}

\noindent 
\begin{equation}
\ident{priority}(x)=\begin{cases}
0 & x=\ident{always}\\
1 & x=\ident{none}\\
2 & x=\ident{nullable}\\
3 & x=\ident{multiple}
\end{cases}
\end{equation}
\noindent
where $a_{i}$ is the applicability of a given term of the \emph{i}th contract
type to be merged.
